---
title: "Homework 01"
pagetitle: "Homework 01"
body-classes: "lecture-page practice-page"
format:
  html:
    embed-resources: true
    html-math-method: mathml
    page-layout: full
    toc: true
    toc-location: body
    toc-title: "On this page"
    toc-depth: 2
---

## Question 1 (Least squares and matrix calculations)

Set the random seed to 43201. Generate $n=100$ observations with $p=5$ predictors according to

$$x_{ij}\overset{\mathrm{iid}}{\sim}N(0,1),
\qquad
\varepsilon_i\overset{\mathrm{iid}}{\sim}N(0,0.7^2),$$

with all predictors and errors generated independently. Let

$$\mathbf X
=
\begin{bmatrix}
\mathbf 1 & \mathbf x_1 & \cdots & \mathbf x_5
\end{bmatrix},
\qquad
\boldsymbol\beta
=
(3,1.4,-0.9,0.7,0,1.1)^{\mathsf T},$$

and generate

$$\mathbf y=\mathbf X\boldsymbol\beta+\boldsymbol\varepsilon.$$

a.  Generate the data and report the dimensions and rank of $\mathbf X$. Calculate the least-squares estimate $\widehat{\boldsymbol\beta}$ using a numerically stable least-squares routine, without explicitly calculating $(\mathbf X^{\mathsf T}\mathbf X)^{-1}$. Compare the estimates with the coefficients used to generate the data.

    ```{r}
    #Set random seed to 43201
    set.seed(43201)

    #Generate a sample of 100 observations with 5 predictors using a normal distribution
    x1 <- rnorm(n = 100, mean = 0, sd = 1)
    x2 <- rnorm(n = 100, mean = 0, sd = 1)
    x3 <- rnorm(n = 100, mean = 0, sd = 1)
    x4 <- rnorm(n = 100, mean = 0, sd = 1)
    x5 <- rnorm(n = 100, mean = 0, sd = 1)

    #Generate a sample of errors for the predictors above
    e <- rnorm(n = 100, mean = 0, sd = 0.7)

    #Create X
    predictors <- data.frame(x1 = x1, x2 = x2, x3 = x3, x4 = x4, x5 = x5)
    X <- cbind(Intercept = 1, as.matrix(predictors))

    #Create beta vector
    beta <- c(3, 1.4, -0.9, 0.7, 0, 1.1)

    #Generate y values and create new dataframe with y
    #y <- 3 + (1.4 * predictors$x1) - (0.9 * predictors$x2) + (0.7 * predictors$x3) + (0 * predictors$x4) + (1.1 * predictors$x5) + e
    y <- (X %*% beta) + e

    data <- predictors
    data <- cbind(data, y = y)

    #Generate least squares regression model
    model <- lm(y ~ x1 + x2 + x3 + x4 + x5, data = data)

    #Find dimensions and rank of X
    dim(X)
    qr(X)$"rank"

    #Calculate the least squares estimate, beta_hat
    beta_hat <- coefficients(model)
    beta_hat
    ```

    The dimensions of X has `r dim(X)[1]` rows and `r dim(X)[2]` columns. The rank of X is `r qr(X)$"rank"`. The estimates for the intercept, x1, x2, x3, and x5 are relatively close to the coefficients used to generate the data. However, the estimate for x4 should be closer to 0.

b.  Calculate the fitted values, the residual vector $\mathbf r=\mathbf y-\mathbf X\widehat{\boldsymbol\beta}$, and the training root mean squared error

$$\operatorname{RMSE}
=
\left(
\frac{1}{n}\sum_{i=1}^{n}r_i^2
\right)^{1/2}.$$

```{r}
#Calculate fitted values
##fitted() generates values for the original data
##predicted() generates values for new data
fitted_values <- fitted(model)

#Calculate residual vector r = y - X * beta_hat
residuals <- resid(model)

#Calculate training root mean squared error
training_RMSE <- ((1 / length(fitted_values)) * sum(residuals ^ 2)) ^ (1 / 2)
training_RMSE
```

c.  Report $\lVert\mathbf X^{\mathsf T}\mathbf r\rVert_\infty$. Explain why this value should be close to zero and what it tells us about the residual vector.

    ```{r}
    #Report the largest value in the vector by
    ##1. Calculate X^T * r
    ##2. Apply the absolute value
    ##3. Use the max() function to find the largest value
    largest_value <- max(abs(t(X) %*% residuals))
    ```

    The maximum norm for the vector X^T^r is `r largest_value`. This value should be close to 0 because the least squares regression strives to minimize errors in its model. In addition, this value tells us that changing any intercepts or predictors would most likely not decrease the residuals.

## Question 2 (Gaussian likelihood for linear regression)

Continue with the data from Question 1. Suppose

$$\mathbf Y\mid\mathbf X
\sim
N_n\left(\mathbf X\boldsymbol\beta,\sigma^2\mathbf I_n\right).$$

The log-likelihood, including the constant term, is

$$\ell(\boldsymbol\beta,\sigma^2)
=
-\frac{n}{2}\log(2\pi\sigma^2)
-\frac{1}{2\sigma^2}
(\mathbf y-\mathbf X\boldsymbol\beta)^{\mathsf T}
(\mathbf y-\mathbf X\boldsymbol\beta).$$

a.  For fixed $\sigma^2$, explain why maximizing $\ell(\boldsymbol\beta,\sigma^2)$ over $\boldsymbol\beta$ is equivalent to minimizing the residual sum of squares. What does this imply about the maximum-likelihood estimate of $\boldsymbol\beta$?

    For a fixed standard deviation, maximizing the log-likelihood over beta is equivalent to minimizing the residual sum of squares because the log-likelihood "measures how well a model explains the data" (<https://www.geeksforgeeks.org/data-science/log-likelihood/>). The second term in the log-likehood formula is the residual sum of squares. Therefore, if the residual sum of squares decreases, then the log-likelihood increases. The maximum-likelihood estimate of beta (coefficients) are the same as the coefficients found under the ordinary least squares method assuming errors are normally distributed (<https://rpubs.com/sharmaar_2024/MLE>).

b.  For fixed $\boldsymbol\beta$, differentiate the log-likelihood with respect to $\sigma^2$ and derive its maximum-likelihood estimate. Calculate this estimate at $\widehat{\boldsymbol\beta}$ and compare it with the unbiased estimate of $\sigma^2$. Explain why the two denominators differ.

    ```{r}
    #Perform a partial derivative on the log-likehood with respect to the standard deviation
    ##Solved the partial derivative by hand
    ##log(a * b) = log(a) + log(b)
    ##(logx)' = 1 / x
    dl_dvar <- expression(-(n / (2 * var)) + (1 / (2 * var ^ 2)) * t((y - X %*% beta)) %*% (y - X %*% beta))

    #Calculate the maximum-likelihood estimate by setting the dl_dvar to 0
    ##Solved for var by hand
    n <- 100
    var <- expression((t((y - X %*% beta)) %*% (y - X %*% beta)) / n)
    MLE <- (t((y - X %*% beta_hat)) %*% (y - X %*% beta_hat)) / n
    MLE

    #Calculate the unbiased estimate of the variance
    ##var = RSS / (n - p)
    ##RSS = residual sum of squares, p = # of estimators including intercept
    p <- 6
    unbiased_var <- (t((y - X %*% beta_hat)) %*% (y - X %*% beta_hat)) / (n - p)
    unbiased_var
    ```

    The maximum likelihood estimate at beta_hat is `r MLE` whereas the unbiased estimate of the variance is `r unbiased_var`. The maximum likelihood estimate is slightly lower than the unbiased estimate. The two denominators differ because the unbiased estimator accounts for the `p` degrees of freedom (<https://stats.stackexchange.com/questions/438438/why-is-the-mle-for-variance-in-single-linear-regression-biased>).

c.  Let

$$\mathbf e_{x_2}=(0,0,1,0,0,0)^{\mathsf T}.$$

Using the variance estimate from part b, plot

$$\ell\left(
\widehat{\boldsymbol\beta}+t\mathbf e_{x_2},
\widehat{\sigma}_{\mathrm{MLE}}^2
\right)$$

over a grid of $t$ values from $-1.25$ to $1.25$. State where the maximum occurs and explain why this agrees with the least-squares result.

```{r}
e_x2 <- c(0, 0, 1, 0, 0, 0)

#Generate a range of t values from -1.25 to 1.25
t <- seq(from = -1.25, to = 1.25, length.out = 100)

#Create a function to evaluate the log-likelihood function
log_likelihood <- function(t_value){
  beta_hat2 <- beta_hat + t_value * e_x2
  RSS <- (t((y - X %*% beta_hat2)) %*% (y - X %*% beta_hat2))
  response <- -(n / 2) * log(2 * pi * MLE) - (1 / (2* MLE)) * RSS
  }

#Plug in every t value into the function and solve
response_vector <- c()
for (i in t){
  response <- as.vector(log_likelihood(i))
  response_vector <- append(response_vector, response)
}

#Plot over a grid of t values from -1.25 to 1.25
plot(x = t, y = response_vector,
     main = "Log-Likelihood of t between -1.25 and 1.25",
     ylab = "response")
```

The maximum occurs at t = 0. This agrees with the least-squares result because at t = 0, the term $t\mathbf e_{x_2}$ becomes zero. Therefore,

$$\widehat{\boldsymbol\beta}=\boldsymbol\beta$$

Under constant variance, the log-likelihood expression becomes

$$\ell(\widehat{\boldsymbol\beta},\widehat{\sigma}_{\mathrm{MLE}}^2)
=
-\frac{n}{2}\log(2\pi\widehat{\sigma}_{\mathrm{MLE}}^2)
-\frac{1}{2\widehat{\sigma}_{\mathrm{MLE}}^2}
(\mathbf y-\mathbf X\widehat{\boldsymbol\beta})^{\mathsf T}
(\mathbf y-\mathbf X\widehat{\boldsymbol\beta}).$$

where the rightmost term is the residual sum of squares. When the residual sum of squares decreases, the log-likelihood increases.

## Question 3 (Starting values in nonconvex optimization)

Consider the function

$$f(x)=\exp(1.5x)-3(x+6)^2-0.05x^3,$$

with derivative

$$f'(x)=1.5\exp(1.5x)-6(x+6)-0.15x^2.$$

a.  Plot $f(x)$ over the interval $[-40,7]$. Based on the plot, describe the important features of the objective function that may affect numerical optimization.

    Source: <https://stackoverflow.com/questions/26091323/how-to-plot-a-function-curve-in-r>

    ```{r}
    curve(exp(1.5 * x) - 3 * (x + 6)^2 - 0.05 * x^3, from = -40, to = 7)
    ```

    The curve gradually increases from -40 to 0, but it suddenly increases to infinity when approaching 7. This sudden increase to infinity may affect numerical optimization because there is no definitive "low point" on this graph.

b.  Use BFGS to minimize $f(x)$ twice, starting at $x^{(0)}=-15$ and $x^{(0)}=0$. Supply $f'(x)$ to the optimizer. For each run, report the final value of $x$, the final objective value, $|f'(x)|$, and whether the optimizer reported convergence.

    ```{r}
    #BFGS = Broyden-Fletcher-Goldfarb-Shanno
    ##Set method to BFGS in optim() function

    #Define function and its first derivative
    q3_fcn <- function(x){
      exp(1.5 * x) - 3 * (x + 6)^2 - 0.05 * x^3
      }
    q3_first_derive <- function(x){
      1.5 * exp(1.5 * x) - 6 * (x + 6) - 0.15 * x^2
    }

    #Use BFGS to minimize f(x) twice at x = -15 and x = 0
    optimize_15 <- optim(par = -15, fn = q3_fcn, gr = q3_first_derive, method = "BFGS")
    optimize_0 <- optim(par = 0, fn = q3_fcn, gr = q3_first_derive, method = "BFGS")
    ```

    At $x^{(0)}=-15$, the final value of $x$ is `optimize_15$par`, the final objective value, $|f'(x)|$, is `r optimize_15$value`, and the optimizer reported a convergence of `r optimize_15$convergence`.

    At $x^{(0)}=0$, the final value of $x$ is `r optimize_0$par`, the final objective value, $|f'(x)|$ is `r optimize_0$value`, and the optimizer reported a convergence of `r optimize_0$convergence`.

c.  Explain why the two runs can converge to different answers. Which run gives the lower objective value? What additional evidence would be needed before claiming that this point is the global minimum?

    The two runs can converge to different answers because they had different starting points. In the graph above, there could be many local minimums around $x^{(0)}=-15$ and $x^{(0)}=0$ that is not visible to us. The `r optim()` function attempts to locate local minimums, not the global minimum, around the given starting point (<https://stackoverflow.com/questions/28546999/why-optimx-in-r-give-different-results-when-changing-the-starting-values>). The run where the function optimizes around $x^{(0)}=-15$ gives the lower objective value. Before claiming that $x^{(0)}=-15$ is the global minimum, the `r optim()` function should be ran at multiple starting points between -40 and 7.

## Question 4 (Nearly collinear predictors)

Continue with $\mathbf X$ and $\mathbf y$ from Question 1. Set the random seed to 43202 and generate

$$u_i\overset{\mathrm{iid}}{\sim}N(0,1),
\qquad i=1,\ldots,n.$$

Define

$$\mathbf x_6=\mathbf x_1+10^{-4}\mathbf u,
\qquad
\mathbf X_+
=
\begin{bmatrix}
\mathbf X & \mathbf x_6
\end{bmatrix},$$

and let

$$\mathbf y^\star=\mathbf y+10^{-3}\mathbf u.$$

Use a numerically stable least-squares routine throughout.

a.  Fit $\mathbf y$ using $\mathbf X$ and $\mathbf X_+$. For each design matrix, report the condition number and training RMSE. For the augmented fit, also report the coefficients of $\mathbf x_1$ and $\mathbf x_6$.

    ```{r}
    #Set random seed to 43202
    set.seed(43202)

    #Generate a sample of 100 observations for u
    u <- rnorm(n = 100, mean = 0, sd = 1)

    #Define x6
    x6 <- x1 + (10^-4) * u

    #Create X_plus by adding x6 to existing X
    X_plus <- cbind(X, x6 = x6)

    #Fit y using X
    ##This has been already done in Question 1
    model

    #Fit y using X_plus
    data_plus <- cbind(data, x6 = x6)
    model2 <- lm(y ~ x1 + x2 + x3 + x4 + x5 + x6, data = data_plus)
    model2

    #Condition Number: "a measure of sensitivity of solving a linear system of equations to variations in the input"
    ##"Small condition numbers correspond to little error amplification. Remember, small condition numbers are good!" (https://cs357.cs.illinois.edu/textbook/notes/condition.html)

    #Find the condition number for X and X_plus
    cond_num_X <- kappa(X)
    cond_num_X_plus <- kappa(X_plus)

    #Calculate the training RMSE for X and X_plus
    ##Calculating training RMSE for X has been done in Question 1

    #Calculate residual vector
    residuals2 <- resid(model2)

    #Calculate training root mean squared error
    training_RMSE2 <- ((1 / n) * sum(residuals2 ^ 2)) ^ (1 / 2)
    ```

    The condition number and training RMSE for $\mathbf X$ is `r cond_num_X` and `r training_RMSE` respectively. The condition number and training RMSE for $\mathbf X_+$ is `r cond_num_X_plus` and `r training_RMSE2` respectively. For the augmented fit, the coefficients of $\mathbf x_1$ and $\mathbf x_6$ are `r coef(model2)["x1"]` and `r coef(model2)["x6"]` respectively.

b.  Fit $\mathbf y^\star$ using $\mathbf X_+$. Report the changes in the coefficients of $\mathbf x_1$ and $\mathbf x_6$. Compare the two fitted-value vectors using their root mean squared difference and maximum absolute difference.

    ```{r}
    #Define y_star
    y_star <- y + (10^-3) * u

    #Add y_star to data
    data_star <- cbind(data_plus, y_star = y_star)

    #Fit y_star using X_plus
    model3 <- lm(y_star ~ x1 + x2 + x3 + x4 + x5 + x6, data = data_star)
    model3

    #Calculate change in coefficients x1 and x6
    change_x1 <- coef(model2)["x1"] - coef(model3)["x1"]
    change_x6 <- coef(model2)["x6"] - coef(model3)["x6"]

    #Calculate root mean squared difference for model of y_star
    residuals3 <- resid(model3)
    training_RMSE3 <- ((1 / n) * sum(residuals3 ^ 2)) ^ (1 / 2)

    #Calculate maximum absolute difference
    ##1. Find the difference between y - y_star
    ##2. Apply the absolute value to the array
    ##3. Find the max value from that array
    max_abs_diff <- max(abs(y - y_star))
    ```

    Between the model fitting $\mathbf y^\star$ using $\mathbf X_+$ to the model fitting $\mathbf y$ using $\mathbf X_+$, an increase of `r change_x1`

    a.  Comparing $\mathbf x_1$ and $\mathbf x_6$ from t

c.  Use the identity

$$\mathbf y^\star-\mathbf y
=
10(\mathbf x_6-\mathbf x_1)$$

to explain why the fitted values change very little while the two coefficients change substantially. What does this example suggest about interpreting separate effects for nearly collinear predictors?

## Question 5 (Data preparation and summary statistics)

The file `data/data-manipulation.csv` is a small constructed data table containing an observation identifier, a group label, two numeric predictors, and a response. Some response values are missing. For analyses involving the response, use only observations with a recorded response. Do not replace a missing response with zero or a group mean.

a.  Read the data and report its dimensions, column types, number of duplicated identifiers, and number of missing values in each column. Create an analysis table containing only observations with a recorded response, and define

$$\texttt{feature\_sum}=\texttt{x1}+\texttt{x2}.$$

b.  For each group, report the number of observations, the mean response, and the mean of `feature_sum`. State clearly which observations these summaries describe.

c.  Sort the analysis table by response from largest to smallest and report the first three rows, including `observation_id`, `group`, `response`, and `feature_sum`. Add code checks verifying that the number of retained rows equals the number of nonmissing responses in the original data, that the analysis table has no missing response values, and that its identifiers are unique.

## Question 6 (Can 39 million Fitbit records represent US adults?)

[Patten et al. (2026)](https://doi.org/10.1038/s41591-026-04352-3) describe Fitbit data from 59,018 participants in the All of Us Research Program. The dataset spans 14 years and contains more than 39 million daily step records. Participants contributed data through one of two routes:

- Bring Your Own Device (BYOD): participants shared data from a Fitbit they already owned.
- Wearables Enhancing All of Us Research (WEAR): invited participants received a Fitbit at no cost.

In the general activity cohort, the BYOD and WEAR groups contained 32,035 and 22,474 participants, respectively. The BYOD value is listed first in each comparison below:

- 77.3% versus 55.1% reported being White;
- 6.1% versus 15.2% reported annual household income between $\$10{,}000$ and $\$25{,}000$; and
- median daily steps were 6,867 versus 5,797.

Suppose the target is the mean of the participant-specific average daily step counts among US adults during the study period, with each adult given equal weight. A researcher writes:

> "This dataset contains more than 39 million daily Fitbit records. Therefore, the average of all recorded step counts should provide an accurate estimate of mean daily activity among US adults."

a.  Are the 39 million daily records independent observations? What characteristics or behaviors could cause some participants to contribute more recorded days than others? Explain how averaging all recorded days would then weight participants unequally.

b.  BYOD participants reported a median of 6,867 daily steps, compared with 5,797 among WEAR participants. Does this comparison show that already owning a Fitbit causes people to walk more? Explain your answer and give at least one plausible alternative explanation based on how participants entered the two groups.

c.  A very large dataset can still produce a biased estimate of a population quantity when the people and observations entering the dataset are selected. Using this study, explain how selection of both participants and recorded days could make the observed data differ from the US adult population. Why might the naive average of all recorded daily step counts fail to estimate the stated target? You do not need to determine the direction of the bias.

### Source

Patten, T., Preble, E. A., Master, H., et al. (2026). [The All of Us Research Program's wearables dataset](https://doi.org/10.1038/s41591-026-04352-3). *Nature Medicine*, 32, 2302-2310.
